Example case study in GMSE

### 2018-05-10 notebook/ms/SI2.Rmd
Minor edits, new read

---
title: 'Use of the gmse_apply function'
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 2)'
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK
[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
linestretch: 1
---

>>>>

---
title: 'Use of the gmse_apply function'
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 2)'
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK
[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
- \usepackage{lineno}
- \linenumbers
linestretch: 1
---

### 2018-05-10 notebook/ms/SI3.Rmd
Minor edits, new read

---
title: 'Example case study in GMSE'
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 3)'
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes: null
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: gmse.bib
biblio-style: apalike
---

@@

Agents in GMSE (managers and users) are goal-oriented, and their behaviour is therefore driven to maximise a particular utility of interest such as a target density of resources, such as animals or trees of conservation interest (managers), or a suitable resource (e.g. animal population size, forest) or landscape harvest size, such as bag size or timber (users). This model feature allows GMSE to evaluate the actions of agents (e.g. managers and users) in the context of their individual objectives, and to therefore quantify the degree to which those objectives are or are not achieved. When the actions of one party clashes with the objectives of another party, the objectives of one might be expressed at the expense of the other, causing conservation conflict [@Redpath2013]. Currently, there is no standard way to measure conservation conflict in a social-ecological system where both the natural resource (e.g. animals, plants) and the people (agents, stakeholders, managers, users etc.) are modelled in a single system, and previous modelling approaches have not meaningfully separated agent objectives from agent actions. We suggest that a starting point to developing a useful metric of conservation conflict is to quantify the deviation of an individual's actions from their objectives  (i.e., of actual actions from desired actions), the former of which is restricted by the actions of other individuals. Here we show how GMSE can be used to evaluate the amount of conflict in a simulated social-ecological system under different management options.

To demonstrate how GMSE can be used to understand conflict in social-ecological systems, we build upon the [example of resource management](https://www.biorxiv.org/content/biorxiv/early/2017/11/17/221432.full.pdf#page=5) in the main text. We consider a protected population of waterfowl that exploits and damages agricultural land and is therefore a source of conservation conflict between those that seek the conservation of waterfowl and those that are concerned with the loss of agriculture [e.g., @Fox2017a; @Mason2017; @Tulloch2017; @Cusack2018]. As in the main text example, the objective of the manager is to keep waterfowl at a target abundance that minmises extinction risks, while the objective of farmers is to maximise agricultural production on their landscape. Here we consider a more complex simulation, with a level of detail that more accurately reflects a scenario that might occur in a real social-ecological system where the manager sets policies that incentivise the user to act in a way that ensures the persistence of the resource. The policies are backed up by a manager budget that can be allocated to set costs of actions and thus incentivise users to follw with the desired actions. The user also has a budget to carry out actions and the cost of the actions is affected by the manager policies and budget. Our objective is not to model the dynamics of a specific system, but to show how GMSE could be parameterised using demographic estimates from empirical studies. We therefore consider an example population in which such estimates are well-reported and readily available.

We parameterise our model using demographic information from the Taiga Bean Goose (*Anser fabalis fabalis*), a managed population that is hunted for recreation in Fennoscandinavia [@Johnson2018]. Taiga Bean Geese can cause agricultural damage [@Johnson2018], which could potentially lead to conflict between farming and management/conservation objectives.

@@

Our goal is not to provide a detailed case study of the Taiga Bean Geese, but rather to demonstrate how such a case study would be possible in GMSE. For simplicity, here we assess conflict using only the `gmse` function to show how parameter values can be set to provide useful results. Novice R users may prefer to run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line. Alternatively, experienced R users may prefer to simulate by [looping time steps](SI2.pdf#page=5) through `gmse_apply`, which allows more flexibility for incorporating custom sub-models and dynamically adjusting parameter values. Where available, we use estimated demographic parameter values from @Johnson2018 and @AEWA2016. Where GMSE parameter values are not available, we use reasonable values or GMSE defaults. To make model inferences for real case studies, we strongly recommend simulating across a range of parameter values when empirical estimates are unavailable, as social-ecological dynamics might be sensitive to these unknown values.

@Johnson2018 recently estimated key demographic parameters of the Taiga Bean Geese from the Central Management Unit, which includes geese that breed in "Northern most Sweden, Northern Norway, Northern and Central Finland, and adjacent North-western parts of Russia, wintering mostly in Southern Sweden and South-east Denmark" [@AEWA2016]. They estimated goose survival under ideal conditions to be ca 0.878; this can be interpreted in our model by setting mortality to `remove_pr = 1 - 0.878`. Similarly, @Johnson2018 estimated mean reproductive rate and carrying capacity to be 0.55 and 93870, respectively, so we set `lambda = 0.275` (for simplicity, we simply use half the mean reproductive rate; GMSE does not currently distinguish female and male individuals) and `res_death_K = 93870`. The global abundance of Taiga Bean Geese in 2009 was ca 63000 [@Fox2010], and ca 35000 in the Central Management Unit [@AEWA2016], which we can take as a starting abundance for our simulations (`RESOURCE_ini = 35000`). And the International Single Species Action Plan has a target population size of ca 70000 in the Central Management Unit [@AEWA2016], which we can use as a management target (`manage_target = 70000`). We simulate social-ecological dynamics over 30 time steps, which could be interpreted as years.

The code below calls `gmse` using the empirically derived parameters for Taiga Bean Geese described above. We also set `manager_budget = 10000` and `user_budget = 10000`. Further, we consider the case of a region in which farmland makes up 60% of all land, with 40% of land being 'public' (`public_land = 0.4`; which might be interpreted as any land in which stakeholders are not, or cannot be, invested in goose presence), and divide the farmland amongst 80 individual farmers (`stakeholders = 80`; `land_ownership = TRUE`). Landscape size is set to default 100 by 100 cells, so each farmer effectively owns 75 cells, which might be interpreted as hectares of land (for instructions on how to more precisely control landscape ownership, see the advanced GMSE options using `gmse_apply`). Because we need both density-dependent (`res_death_K = 93870`) and density-independent (`remove_pr = 0.122`) sources of mortality, we set `res_death_type = 3`. We assume that a single goose decreases agricultural production on a cell by 2% per time step (`res_consume = 0.02`). We further assume that the population is very well-monitored, with observers counting goose numbers on each cell of the landscape in every timestep (`observe_type = 3`) with the ability to observe one landscape cell in every direction (`agent_view = 1`). All other parameter values are set to GMSE defaults.

@@

```{r, eval = TRUE, fig.height = 7, fig.cap = "Results of a GMSE simulation using parameters estimated for Taiga Been Geese Central Management Unit. This example includes 80 farmers whose objective is to maximise their agricultural output, and one manager whose objective is to keep geese at a target abundance, over 30 simulated time steps. Goose locations at the end of the simulation are shown in the upper left panel, while the upper right panel shows the same landscape broken down among the 80 farmers (upper 60% of the landscape in multiple colours), along with non-agricultural land (lower 40% of the landscape in blue). Actual goose abundance is shown in the middle left panel (black solid line), along with its estimate by the manager (blue solid line, hidden underneath the black line). The horizontal red and blue dotted lines show goose carrying capacity and the manager's target for goose abundance, respectively. The organge line shows the total percent of landscape cell (including non-farmed cells) yield, as decreased by geese. The middle right panel shows this yield for each farmer, and for the public land (lower line in blue). The lower left panel shows the cost of culling for farmers, as set by the manager, and the lower right panel shows the total number of culls attempted by farmers over time."}
plot_gmse_results(res = sim_1$resource, obs = sim_1$observation, 
                  land = sim_1$land, agents = sim_1$agents, paras = sim_1$paras, 
                  ACTION = sim_1$action, COST = sim_1$cost);
```

Figure 1 shows the dynamics of goose abundance and agricultural yield, along with how managers react to change in abundance and farmers react to manager policy. In the case of the simulation above, managers quickly set a policy of high cost for culling, which leads to a rise in the goose population and a decrease in crop yield for farmers. After roughly 20 time steps, the goose population rises above the manager target, at which point the manager becomes more permissive of culling and the cost of culling for users therefore declines. In response, users begin to cull geese on their land, and the goose population begins to stabilise around the target of 7000 total geese. We can investigate the conflict between management policy and farmers more directly using the `plot_gmse_effort` function.

@@

Black lines indicate how permissive a manager is toward a particular action on a scale of 0 to 100, while coloured lines indicate how much effort farmers expend on a given action. When black lines are far below coloured lines, we can (cautiously) interpret this as a conflict between management of the goose population and farmer's interest in agricultural production. These time periods represent instances in which the manager is not permissive of a particular action (in this case culling), but farmers continue to expend effort to do the action anyway. In the case of the above simulation of potential conflict between farmers and goose conservation, conflict is highest before time step 20, where the manager is not permissive of culling because the population is below the manager's target. Once the goose population has increased above the manager's target, conflict decreases because the desired culling is permitted by managers to keep the population at a target abundance. It is worth noting that, despite conflict as we define it decreasing, agricultural damage is still relatively high after the target goose population size is achieved (Figure 1). Hence, on a broader scale, conflict might persist around the appropriate target population size rather than what actions are permitted for farmers; currently, this potential aspect of conflict is not modelled, but future versions of GMSE may attempt to incorporate such additional complexity in conflict scenarios.

@@

```{r, eval = TRUE, fig.height = 8, fig.cap = "Results of a GMSE simulation using parameters estimated for Taiga Been Geese Central Management Unit in which scaring is permitted. Simulation output is interpreted as in Figure 1"}
plot_gmse_results(res = sim_2$resource, obs = sim_2$observation, 
                  land = sim_2$land, agents = sim_2$agents, paras = sim_2$paras,
                  ACTION = sim_2$action, COST = sim_2$cost);
```

@@

The GMSE function `gmse` and its graphical user interface counterpart `gmse_gui` offer wide a suite of options for parameterising simulations to fit empirical case studies of conservation interest using default GMSE natural resource, observation, manager, and user sub-models. Future development of these sub-models might usefully incorporate additional details relevant to specific case studies, such population structure, multiple natural resource and user types, or different manager policy and user action possibilities. [Requests for new features](https://github.com/bradduthie/gmse/issues) and contributions to GMSE can be made through [GitHub](https://github.com/bradduthie/gmse). Additionally, where entirely different types of sub-models are required, the `gmse_apply` function [can be used](https://bradduthie.github.io/gmse_SI2.pdf) to more flexibly simulate social-ecological systems.

>>>>

---
title: 'Example case study in GMSE'
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 3)'
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
- \usepackage{lineno}
- \linenumbers
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: gmse.bib
biblio-style: apalike
---

@@

Agents in GMSE (managers and users) are goal-oriented, and their behaviour is therefore driven to maximise a particular utility of interest such as a target density of resources. For managers and users, this could include animals or trees of conservation interest. For users, it could additionally include a landscape harvest size such as bag size or timber. This model feature allows GMSE to evaluate the actions of agents in the context of their individual objectives, and to therefore quantify the degree to which those objectives are or are not achieved. When the actions of one party clashes with the objectives of another party, the objectives of one might be expressed at the expense of the other, causing conservation conflict [@Redpath2013]. Currently, there is no standard way to measure conservation conflict in a social-ecological system where both the natural resource (e.g., animals, plants, or non-biological resources) and the people (e.g., stakeholders, managers, etc.) are modelled in a single system, and previous modelling approaches have not meaningfully separated agent objectives from agent actions. We suggest that a starting point to developing a useful metric of conservation conflict is to quantify the deviation of an individual's actions from their objectives  (i.e., of actual actions from desired actions), the former of which is restricted by the actions of other individuals. Here we show how GMSE can be used to evaluate the amount of conflict in a simulated social-ecological system under different management options.

To demonstrate how GMSE can be used to understand conflict in social-ecological systems, we build upon the [example of resource management](https://www.biorxiv.org/content/biorxiv/early/2017/11/17/221432.full.pdf#page=5) in the main text. We consider a protected population of waterfowl that exploits and damages agricultural land and is therefore a source of conservation conflict between those that seek the conservation of waterfowl and those that are concerned with the loss of agriculture [e.g., @Fox2017a; @Mason2017; @Tulloch2017; @Cusack2018]. As in the main text example, the objective of the manager is to keep waterfowl at a target abundance that minimises extinction risks, while the objective of farmers is to maximise agricultural production on their landscape. Here we consider a more complex simulation, with a level of detail that more accurately reflects a scenario that might occur in a real social-ecological system where the manager sets policies that incentivise the user to act in a way that ensures the persistence of the resource. The policies are backed up by a manager budget that can be allocated to set costs of actions and thus incentivise users to perform the desired actions. The user also has a budget to carry out actions, and the cost of the user actions is affected by the manager policies and budget. Our objective is not to model the dynamics of a specific system, but to show how GMSE could be parameterised using demographic estimates from empirical studies. We therefore consider an example population in which such estimates are well-reported and readily available.

We parameterise our model using demographic information from the Taiga Bean Goose (*Anser fabalis fabalis*), a managed population that is hunted for recreation in Fennoscandinavia [@Johnson2018]. Taiga Bean Geese can cause agricultural damage [@Johnson2018], which could potentially lead to conflict between farming and management or conservation objectives.

@@

Our goal is not to provide a detailed case study of the Taiga Bean Geese, but rather to demonstrate how such a case study would be possible in GMSE. For simplicity, here we assess conflict using only the `gmse` function to show how parameter values can be set to provide useful results. Novice R users may prefer to run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line. Alternatively, experienced R users may prefer to simulate by [looping time steps](SI2.pdf#page=5) through `gmse_apply`, which allows more flexibility for incorporating custom sub-models and dynamically adjusting parameter values. Where available, we use estimated demographic parameter values from @Johnson2018 and @AEWA2016. Where GMSE parameter values are not available, we use reasonable values or GMSE defaults. To make model inferences for real case studies, we strongly recommend [replicating simulations](SI6.pdf) and simulating across a range of parameter values when empirical estimates are unavailable, as social-ecological dynamics might be sensitive to these unknown values.

@Johnson2018 recently estimated key demographic parameters of the Taiga Bean Geese from the Central Management Unit, which includes geese that breed in "Northern most Sweden, Northern Norway, Northern and Central Finland, and adjacent North-western parts of Russia, wintering mostly in Southern Sweden and South-east Denmark" [@AEWA2016]. They estimated goose survival under ideal conditions to be ca 0.878; this can be interpreted in our model by setting mortality to `remove_pr = 1 - 0.878`. Similarly, @Johnson2018 estimated mean reproductive rate and carrying capacity to be 0.55 and 93870, respectively, so we set `lambda = 0.275` (for simplicity, we simply use half the mean reproductive rate; GMSE does not currently distinguish female and male individuals) and `res_death_K = 93870`. The global abundance of Taiga Bean Geese in 2009 was ca 63000 [@Fox2010], with ca 35000 in the Central Management Unit [@AEWA2016], which we can take as a starting abundance for our simulations (`RESOURCE_ini = 35000`). The International Single Species Action Plan has a target population size of ca 70000 in the Central Management Unit [@AEWA2016], which we can use as a management target (`manage_target = 70000`). We simulate social-ecological dynamics over 30 time steps, which could be interpreted as years.

The code below calls `gmse` using the empirically derived parameters for Taiga Bean Geese described above. We also set `manager_budget = 10000` and `user_budget = 10000`. Further, we consider the case of a region in which farmland makes up 60% of all land, with 40% of land being 'public' (`public_land = 0.4`; which might be interpreted as any land in which stakeholders are not, or cannot be, invested in goose presence), and divide the farmland amongst 80 individual farmers (`stakeholders = 80`; `land_ownership = TRUE`). Landscape size is set to default 100 by 100 cells, so each farmer effectively owns 75 cells, which might be interpreted as hectares of land (for instructions on how to more precisely control landscape ownership, see the [advanced GMSE options](SI4.pdf) using `gmse_apply`). Because we need both density-dependent (`res_death_K = 93870`) and density-independent (`remove_pr = 0.122`) sources of mortality, we set `res_death_type = 3`. We assume that a single goose decreases agricultural production on a cell by 2% per time step (`res_consume = 0.02`). We further assume that the population is very well-monitored, with observers counting goose numbers on each cell of the landscape in every timestep (`observe_type = 3`) with the ability to observe one landscape cell in every direction (`agent_view = 1`). All other parameter values are set to GMSE defaults.

@@

```{r, eval = TRUE, fig.height = 7, fig.cap = "Results of a GMSE simulation using parameters estimated for Taiga Been Geese Central Management Unit. This example includes 80 farmers whose objective is to maximise their agricultural output, and one manager whose objective is to keep geese at a target abundance, over 30 simulated time steps. Goose locations at the end of the simulation are shown in the upper left panel, while the upper right panel shows the same landscape broken down among the 80 farmers (upper 60% of the landscape in multiple colours), along with non-agricultural land (lower 40% of the landscape in blue). Actual goose abundance is shown in the middle left panel (black solid line), along with its estimate by the manager (blue solid line, hidden underneath the black line). The horizontal red and blue dotted lines show goose carrying capacity and the manager's target for goose abundance, respectively. The orange line shows the total percent of landscape cell (including non-farmed cells) yield, as decreased by geese. The middle right panel shows this yield for each farmer, and for the public land (lower line in blue). The lower left panel shows the cost of culling for farmers, as set by the manager, and the lower right panel shows the total number of culls attempted by farmers over time."}
plot_gmse_results(res = sim_1$resource, obs = sim_1$observation, 
                  land = sim_1$land, agents = sim_1$agents, paras = sim_1$paras, 
                  ACTION = sim_1$action, COST = sim_1$cost);
```

Figure 1 shows the dynamics of goose abundance and agricultural yield, along with how managers react to change in abundance and farmers react to manager policy. In the case of the simulation above, managers quickly set a policy of high cost for culling, which leads to a rise in the goose population and a decrease in crop yield for farmers. After roughly 20 time steps, the goose population rises above the manager target, at which point the manager becomes more permissive of culling and the cost of culling for users therefore declines. In response, users begin to cull geese on their land, and the goose population begins to stabilise around the target of 7000 total geese. We can investigate the conflict between management policy and farmers more directly using the `plot_gmse_effort` function (Figure 2).

@@

Black lines in Figure 2 indicate how permissive a manager is toward a particular action on a scale of 0 to 100, while coloured lines indicate how much effort farmers expend on a given action. When black lines are far below coloured lines, we can (cautiously) interpret this as a conflict between management of the goose population and farmer's interest in agricultural production. These time periods represent instances in which the manager is not permissive of a particular action (in this case culling), but farmers continue to expend effort to do the action anyway. In the case of the above simulation of potential conflict between farmers and goose conservation, conflict is highest before time step 20, where the manager is not permissive of culling because the population is below the manager's target. Once the goose population has increased above the manager's target, conflict decreases because the desired culling is permitted by managers to keep the population at a target abundance. It is worth noting that, despite conflict as we define it decreasing, agricultural damage is still relatively high after the target goose population size is achieved (Figure 1). Hence, on a broader scale, conflict might persist around the appropriate target population size rather than what actions are permitted for farmers; currently, this potential aspect of conflict is not modelled, but future versions of GMSE may attempt to incorporate such additional complexity in conflict scenarios.

@@

```{r, eval = TRUE, fig.height = 8, fig.cap = "Results of a GMSE simulation using parameters estimated for Taiga Been Geese Central Management Unit in which scaring is permitted. Simulation output is interpreted as in Figure 1."}
plot_gmse_results(res = sim_2$resource, obs = sim_2$observation, 
                  land = sim_2$land, agents = sim_2$agents, paras = sim_2$paras,
                  ACTION = sim_2$action, COST = sim_2$cost);
```

@@

The GMSE function `gmse` and its graphical user interface counterpart `gmse_gui` offer wide a suite of options for parameterising simulations to fit empirical case studies of conservation interest using default GMSE natural resource, observation, manager, and user sub-models. Future development of these sub-models might usefully incorporate additional details relevant to specific case studies, such population structure, multiple natural resource and user types, or different manager policy and user action possibilities. [Requests for new features](https://github.com/bradduthie/gmse/issues) and contributions to GMSE can be made through [GitHub](https://github.com/bradduthie/gmse). Additionally, where entirely different types of sub-models are required, the `gmse_apply` function [can be used](https://bradduthie.github.io/gmse_SI2.pdf) to more flexibly simulate social-ecological systems. In [SI4](SI4.pdf), we show an example of this using the same Taiga Bean Geese case study that we did here.

\clearpage
